Julianna's algal metrics wrangling code

### 2024-07-30 julianna_algalMetric_wrangling.Rmd
Updated wrangling/metric calculating code

```{r}
# algal data
algae <- read_csv(here("data", "algaeonly_cleaned_colnames.csv")) %>% 
  mutate(island_side = case_when(site == "lter_1" ~ "north",
                                 site == "lter_2" ~ "north",
                                 site == "lter_3" ~ "east",
                                 site == "lter_4" ~ "east",
                                 site == "lter_5" ~ "west",
                                 site == "lter_6" ~ "west",
                                 TRUE ~ "UdINGLEbUTT"))
```

>>>>

```{r}
# algal cover 
algae <- read_csv(here("data", "algaeonly_cleaned_colnames.csv")) %>% 
  mutate(island_side = case_when(site == "lter_1" ~ "north",
                                 site == "lter_2" ~ "north",
                                 site == "lter_3" ~ "east",
                                 site == "lter_4" ~ "east",
                                 site == "lter_5" ~ "west",
                                 site == "lter_6" ~ "west",
                                 TRUE ~ "UdINGLEbUTT"))

# sand cover
sand <- read_csv(here("data", "allbenthicdata_cleaned_colnames.csv")) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, sand, coral)
```

# Calculate algal alpha diversity

## Make matrices

Make a spp matrix to use with `vegan`

```{r make species matrix}
# matrix of just species and their relative percent covers in the same order as the algae tibble
algae %>% 
  select(-year, -location, -site, -island_side, -habitat, -site_habitat, -transect, -quadrat) -> sppMat
```

```{r make a metadata matrix in the same order}
# get ranges to calculate over (this is just for indexing purposes in the next bit of code)
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# metadata matrix
algae %>% 
  # bring in sand data
  full_join(sand) %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg]), # percent cover of algae
         AlgCov_percHard = case_when(
           sand == 0 ~ AlgCov_perc, # since we can't divide by a zero
           TRUE ~ AlgCov_perc/sand 
           )) %>% # percent cover of hard substrate
   # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, "_", hab_year)) %>% 
  select(year, location, site, habitat, island_side, site_habitat, 
         hab_year, island_side_hab_year, transect, quadrat, AlgCov_perc, 
         AlgCov_percHard, sand, coral)  -> sppMeta
```

## Calculate diversity metrics

```{r get richness and Shannon diversity}
sppMeta %>% 
  # these functions use the vegan package
  mutate(Richness = specnumber(sppMat),
         Shannon = diversity(sppMat, "shannon")) -> algDiv
```

# Calculate algal beta diversity

## Distance matrix: habitat-year

We get an error if we run `vegdist()` with quadrats that have no algae. We can see there are 672 of these that we need to filter out:

```{r look for zeros}
sppMat %>% 
  mutate(AlgCov_perc = rowSums(across(where(is.numeric)))) %>% 
  filter(AlgCov_perc == 0) %>% dim() # 672 with no algae at all
```

Now we can exclude quadrats with no algae and make a species matrix again

```{r get spp mat df}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix where 0 observations are filtered out
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0
```

```{r metadata matix}
# matching metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year)) -> sppMeta0
```

Now we'll calculate this by habitat because we're compute limited

### Calculate fringing

```{r get fringe mat}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  filter(habitat == "fringing") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0fring

# metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
   filter(habitat == "fringing") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0fring
```

Calculate Bray Curtis distances

```{r calculate distances}
brayMatFring <- vegdist(sppMat0fring, "bray") # 1 is maximum dissimilarity
```

Calculate dispersion (this is using Noam's code)

```{r calculate beta dispersion}
yrBetaDisperFring <- vegan::betadisper(brayMatFring, sppMeta0fring$year) # warning: some squared distances are negative and changed to zero (ask Jamie about this??)
```

### Backreef

```{r get backreef df}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  filter(habitat == "backreef") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0back

# metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
   filter(habitat == "backreef") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0back
```

Calculate Bray Curtis distances

```{r calculate distance}
brayMatBack <- vegdist(sppMat0back, "bray") # 1 is maximum dissimilarity
```

Calculate beta disperion

```{r calculate beta dispersion}
yrBetaDisperBack <- vegan::betadisper(brayMatBack, sppMeta0back$year) 
```

### Outer 10

```{r get outer 10 df}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  filter(habitat == "outer_10") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0outer10

# metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
   filter(habitat == "outer_10") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0outer10
```

Calculate Bray Curtis distances

```{r calculate distance}
brayMatOuter10 <- vegdist(sppMat0outer10, "bray") # 1 is maximum dissimilarity
```

Calculate beta disperion

```{r calculate beta dispersion}
yrBetaDisperOuter10 <- vegan::betadisper(brayMatOuter10, sppMeta0outer10$year) # error: some squared distances are negative and changed to zero
```

### Outer 17

```{r get outer 17 df}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

# spp matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  filter(habitat == "outer_17") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0outer17

# metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
   filter(habitat == "outer_17") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0outer17
```

Calculate Bray Curtis distances

```{r calculate distance}
brayMatOuter17 <- vegdist(sppMat0outer17, "bray") # 1 is maximum dissimilarity
```

Calculate beta disperion

```{r calculate beta dispersion}
yrBetaDisperOuter17 <- vegan::betadisper(brayMatOuter17, sppMeta0outer17$year) # error: some squared distances are negative and changed to zero
```

## Join together

Attach back to metadata

```{r join distances with metadata}
sppMeta0fring %>% 
  mutate(distance_centriod = yrBetaDisperFring$distances) -> fringeDist

sppMeta0back %>% 
  mutate(distance_centriod = yrBetaDisperBack$distances) -> backDist

sppMeta0outer10 %>% 
  mutate(distance_centriod = yrBetaDisperOuter10$distances) -> outer10Dist

sppMeta0outer17 %>% 
  mutate(distance_centriod = yrBetaDisperOuter17$distances) -> outer17Dist
```

```{r join together}
# make sure this looks good
colnames(fringeDist) == colnames(backDist) 
colnames(backDist) == colnames(outer10Dist) 
colnames(outer10Dist) == colnames(outer17Dist)

# join together and tidy up
betaDHabYr <- rbind(fringeDist, backDist, outer10Dist, outer17Dist) %>% 
  mutate(hab_yr_distcentroid = distance_centriod) %>% 
  select(year, location, site, habitat, transect, quadrat, hab_yr_distcentroid)
```

### Save dispersions

```{r}
# save(yrBetaDisperFring, file = here("data/fringingyrdispersion.Rdata"))
# save(yrBetaDisperBack, file = here("data/backyrdispersion.Rdata"))
# save(yrBetaDisperOuter10, file = here("data/outer10yrdispersion.Rdata"))
# save(yrBetaDisperOuter17, file = here("data/outer17yrdispersion.Rdata"))
```

### Load dispersions

```{r}
load(here("data/fringingyrdispersion.Rdata"))
load(here("data/backyrdispersion.Rdata"))
load(here("data/outer10yrdispersion.Rdata"))
load(here("data/outer17yrdispersion.Rdata"))
```

# Join all quadrat-level datasets together

```{r join}
algDiv %>% 
  full_join(betaDHabYr) %>% 
  # follow a cleaner naming convention
  rename(algal_perc_cov = AlgCov_perc,
         algal_perc_cov_hard = AlgCov_percHard,
         algal_richness = Richness,
         algal_Shann = Shannon) %>% 
  # get just the columns we need
  select(-site_habitat, -hab_year, -island_side_hab_year) -> quadData
```

```{r save}
# write_csv(quadData, here("algal_metrics_quadrat.csv"))
```

# Site/habitat/year-level mean

We'll also average these data at the site/habitat/year-level to be more compatible with other data types. This will be done at the *quadrat-level*, meaning we won't take the transect-level summary first, since they aren't meant to be factors in a statistical design

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(mean_coral_perc_cov = mean(coral),
            mean_alg_perc_cov = mean(algal_perc_cov),
            mean_alg_perc_cov_hard = mean(algal_perc_cov_hard),
            mean_alg_richness = mean(algal_richness),
            mean_alg_Shann = mean(algal_Shann),
            mean_dist_centroid_habyear = mean(hab_yr_distcentroid, na.rm = TRUE)
            ) -> siteData
```

```{r save dataframe}
# write_csv(siteData, here("algal_metrics_site.csv"))
```

# All time metrics

## Calculate the coefficient of variation

This is basically just copying/pasting Noam's code

```{r cv function}
# coefficient of variation
CV <- function(x){
  return(sd(x,na.rm=T)/mean(x,na.rm=T))
}
```

Apply it when we summarize

```{r calculate CV and summarize}
quadData %>% 
  # don't include year this time
  group_by(site, habitat, island_side) %>%
  summarize(
    mean_coral_perc = mean(coral),
    mean_alg_richness = mean(algal_richness), 
    richness_se = sd(algal_richness)/sqrt(n()),
    richness_cv = CV(algal_richness),
    mean_Shann = mean(algal_Shann), 
    Shann_se = sd(algal_Shann)/sqrt(n()),
    Shann_cv = CV(algal_Shann),
    # for total algal cover
    mean_alg_cov = mean(algal_perc_cov),
    alg_cov_se = sd(algal_perc_cov)/sqrt(n()),
    cv_alg_cov = CV(algal_perc_cov),
    # and for % of hard substrate
    mean_alg_cov_hard = mean(algal_perc_cov_hard),
    alg_cov_hard_se = sd(algal_perc_cov_hard)/sqrt(n()),
    cv_alg_cov_hard = CV(algal_perc_cov_hard),
    ) -> alltimeCV
```

```{r save}
# write_csv(alltimeCV, here(here("algal_metrics_all_time.csv")))
```

- Noam: Add **synchrony** to this dataset + any other relevant metrics

s
